The Dark Matter Distribution of the Local Group inferred from Cosmological Simulations: I. The Halo Masses of M31, M33 and the Milky Way

### 2015-04-10 doc/thesis/main.tex
Updated abstract, intro

\usepackage{aas_macros}
\usepackage{natbib}
\bibliographystyle{apj}

@@

We study the local group of galaxies in its cosmological context, by matching
dark matter halos in the \consuelo N-body simulation with kinematic
observations of M31 and M33. Treating the \consuelo halo  catalog as a list of
samples drawn from the cosmological prior PDF for the local group properties,
we select local group analogs whose isolation matches our own, and then
compute importances for the samples as combined likelihoods for the
observations. 
%
From the ``observed'' galactocentric distance, radial velocity and tangential 
velocity of M31 alone, we infer the mass of the local group to be $\MLG =
(\MMW + \MEI) = (\MPAIRestimate^{+\MPAIRerrorplus}_{-\MPAIRerrorminus}) \times
10^{12} \Msun$. This is ... than the mass that has been estimated from the
timing argument: we show that for our matched analogs the timing argument mass
systematically ... the local group mass. 
%
Transforming the ``observed'' galactocentric distance, radial velocity and
tangential  velocity of M33 into the M31 rest-frame, and accounting for
covariance between these measurements, we also infer the mass of the M31--M33
system to be  $(\MEI + \MEE) =
(\MbAestimate^{+\MbAerrorplus}_{-\MbAerrorminus}) \times 10^{12} \Msun$.
%
Combining the kinematic constraints from both M31 and M33,  we infer a local
group mass of $\MLG = (\MMW + \MEI) =
(\MTRIPLETestimate^{+\MTRIPLETerrorplus}_{-\MTRIPLETerrorminus}) \times
10^{12} \Msun$, and are able to estimate the halo masses of the three galaxies
individually, finding these to be $\MEI =
(\MEIestimate^{+\MEIerrorplus}_{-\MEIerrorminus}) \times 10^{12} \Msun$, $\MEE
= (\MEEestimate^{+\MEEerrorplus}_{-\MEEerrorminus}) \times 10^{12} \Msun$, 
and $\MMW = (\MMWestimate^{+\MMWerrorplus}_{-\MMWerrorminus}) \times 10^{12}
\Msun$. 
%
This Milky Way halo mass estimate is in xxxagreement with previous estimates
from stellar and dwarf satellite kinematics.

@@

Our goal is to map out the dark matter distribution in the Local Group
by combining observational data on the  positions and velocities of the
three largest LG members with our understanding of structure formation
in $\Lambda$CDM universes, as detailed in halo catalogs extracted from
cosmological N-body simulations. We treat halo catalogs as containing
samples drawn from a very informative, highly correlated prior PDF for
halo properties, and then  importance sample them with likelihood
functions based on the observed kinematics of the particular system
under study. In \citet[][hereafter B11]{Bus++11} we presented a new
measurement of the Milky Way halo mass, made in this way using the
\bolshoi simulation of \citet{Bolshoi} and the distances to and speeds
of the Magellanic Clouds. In this work, we use the lower resolution but
higher volume \consuelo simulations, and constrain them with the radial
velocities, proper motions and distances of M31 and M33, as well as our
previous MW mass estimate.

@@

In short, we seek answers to the following questions:
\begin{itemize}
\item What is the mass of dark matter in the Local Group?
\item How accurate is the Timing Argument for the Local Group mass?
\item What are the masses of the dark matter halos of M31, M33 and the Milky
Way?
\end{itemize}

This paper is structured as follows. In \Sref{sec:method} we review the
halo inference methodology, and introduce the \consuelo N-body
simulation suite we use. Then in \Sref{sec:data} we summarise the
observational constraints we place on our model halo population during
the importance sampling process: this includes the isolation criteria we
use in the initial halo selection step. In \Sref{Sec:results} we present
our results on the mass of the Local Group and its members, and then
discuss those results in the context of previous work in
\Sref{sec:discuss}. We conclude in \Sref{sec:conclude}.

@@

In this section we describe the coordinate systems used in this paper and present the transformations between them.  As mentioned in the introduction, this project uses the M31 galactocentric dynamics of the MW-M31-M33 system to define a LG analog.  Specifically, we use the parameters $\distance$, $\vrad$, and $\vtan$ measured from M31 to MW and M33 respectively.  The LG analog sample is drawn using the "observed" values of these parameters.  However, the true observed values of these parameters are of course geocentric and measured in equatorial coordinates of right ascension ($\alpha$) and declination ($\delta$).  Therefore, it is necessary to be able to transform from this system to the M31 galactocentric coordinate system.  This involves many intermediate coordinate systems and transformations which will be described in the remainder of this section.
\subsection{Coordinate System Definitions}
\label{sec:c_sys_defs}
Three types of coordinate systems are used in this work: Geocentric Equatorial, Heliocentric Galactic, and Galactocentric.  Before we describe the mathematical transformations between these systems, it is important to thoroughly understand the definitions of the systems. 
\par The equatorial system is centered on the Earth and based on the concept of a celestial sphere (ie all stars are projected onto a sphere of infinite radius centered on the Earth). Two angles characterize the positions of objects on the celestial sphere: right ascension and declination. Right ascension, $\alpha$, is the angle measured eastwards along the celestial equator between the vernal equinox and the hour circle passing through an object.  Declination, $\delta$, is the angular distance of an object above or below the celestial equator.  
\par The heliocentric galactic coordinate system (often ambiguously referred to as galactic coordinates), is a different spherical coordinate system centered at the Sun.  The heliocentric fundamental plane is aligned with the galactic plane, and the primary direction is along the line connecting the sun and the galactic center.  Positions are measured in galactic latitude and longitude. Latitude, $b$, measures angular distance of an object perpendicular to the galactic plane, with positive angles to the north galactic pole. Longitude, $l$, is the angular distance measured eastward along the galactic equator from the galactic center.
\par The galactocentric coordinate system used in this paper is the same as the one used by van der Marel \citep{VdM08}.  The $Z$ axis points towards the galactic north pole, the $Y$ axis is oriented in the direction of the sun's rotation about the galactic center, and the $X$ axis points from the sun towards the galactic center. This is a standard, right-handed coordinate system.

@@

\todo{Phil}{Write this section!}

@@

\subsection{The \consuelo Simulation}
\label{sec:sampling}

\todo{Mike}{Write this section!}

@@

\subsection{The Local Group prior}
\label{sec:sampling}

\todo{Mike}{Write this section!}

@@

@@

>>>>

\usepackage{aas_macros}
\usepackage{natbib}
%\usepackage{siunitx}
\bibliographystyle{apj}

@@

We study the local group of galaxies by identifying analogous systems in the \consuelo N-body simulation.  The \consuelo halo catalogs are treated as a large sample drawn from the cosmological prior PDF, and local group analogs are identified by matching the observed kinematics of the system.  Weights are calculated for the \consuelo sample by finding the likelihoods for the simulated kinematic properties compared to observations.  We study the effect of including the existence and kinematics of M33 on the posterior distributions of the local group mass by using three types of samples: from only the "observed" galactocentric distance, radial velocity, and tangential velocity of M31, we infer the mass of the local group to be $\MLG =
(\MMW + \MEI) = (\MPAIRestimate^{+\MPAIRerrorplus}_{-\MPAIRerrorminus}) \times
10^{12} \Msun$. Including the only the existence of M33 as a condition for identifying local group analogs results in a different inference of the mass, $\MLG =
(\MMW + \MEI) = (\MPAIRestimate^{+\MPAIRerrorplus}_{-\MPAIRerrorminus}) \times
10^{12} \Msun$. Finally including the kinematics of M33 we infer the mass of the local group to be $\MLG =
(\MMW + \MEI) = (\MPAIRestimate^{+\MPAIRerrorplus}_{-\MPAIRerrorminus}) \times
10^{12} \Msun$.  This is ... than the mass that has been estimated from the
timing argument: we show that for our matched analogs the timing argument mass
systematically ... the local group mass.  We are also able to estimate the halo masses of the three galaxies
individually, finding these to be $\MEI =
(\MEIestimate^{+\MEIerrorplus}_{-\MEIerrorminus}) \times 10^{12} \Msun$, $\MEE
= (\MEEestimate^{+\MEEerrorplus}_{-\MEEerrorminus}) \times 10^{12} \Msun$, 
and $\MMW = (\MMWestimate^{+\MMWerrorplus}_{-\MMWerrorminus}) \times 10^{12}
\Msun$. 
%
This Milky Way halo mass estimate is in xxxagreement with previous estimates
from stellar and dwarf satellite kinematics.

@@

Our goal is to tightly constrain the mass of the Local Group
by combining observational data on the  positions and velocities of the
three largest LG members with our understanding of structure formation
in $\Lambda$CDM universes, as detailed in halo catalogs extracted from
cosmological N-body simulations. We treat halo catalogs as containing
samples drawn from a very informative, highly correlated prior PDF for
halo properties, and then  calculate weighted posterior PDF's using likelihood
functions based on the observed kinematics of the particular system
under study. In \citet[][hereafter B11]{Bus++11} we presented a new
measurement of the Milky Way halo mass, made in this way using the
\bolshoi simulation of \citet{Bolshoi} and the distances to and speeds
of the Magellanic Clouds. In this work, we use the lower resolution but
higher volume \consuelo simulations, and constrain them with the radial
velocities, proper motions and distances of M31 and M33, as well as our
previous MW mass estimate.

@@

In short, we seek answers to the following questions:
\begin{itemize}
\item What is the mass of dark matter in the Local Group?
\item How accurate is the Timing Argument for the Local Group mass?
\item How do the existence and kinematics of M33 impact the Local Group mass?
\item What are the masses of the dark matter halos of M31, M33 and the Milky
Way?
\item How sensitive is this method to the isolation criteria used to define an LG-analog?
\end{itemize}

This paper is structured as follows. In \Sref{sec:coord} we review the coordinate systems and transformations used in this project.  In \Sref{sec:method} we review the
halo inference methodology, and introduce the \consuelo N-body
simulation suite we use. Then in \Sref{sec:data} we summarise the
observational constraints we place on our model halo population during
the importance sampling process: this includes the isolation criteria we
use in the initial halo selection step. In \Sref{Sec:results} we present
our results on the mass of the Local Group and its members, and then
discuss those results in the context of previous work in
\Sref{sec:discuss}. We conclude in \Sref{sec:conclude}.

@@

In this section we describe the coordinate systems used in this paper and present the transformations between them.  As mentioned in the introduction, this project uses the M31 galactocentric dynamics of the MW-M31-M33 system to define a LG analog.  Specifically, we use the parameters $\distance$, $\vrad$, and $\vtan$ measured from M31 to MW and M33 respectively.  The LG analog sample is drawn using the "observed" values of these parameters.  However, the true observed values of these parameters are of course geocentric and measured in equatorial coordinates of right ascension ($\alpha$) and declination ($\delta$).  Therefore, it is necessary to be able to transform from this system to the M31 galactocentric coordinate system.  This involves many intermediate coordinate systems and transformations which will be described in the remainder of this section.
\subsection{Coordinate System Definitions}
\label{sec:c_sys_defs}
Three types of coordinate systems are used in this work: Geocentric Equatorial, Heliocentric Galactic, and Galactocentric.  Before we describe the mathematical transformations between these systems, it is important to thoroughly understand the definitions of the systems. 
\par The equatorial system is centered on the Earth and based on the concept of a celestial sphere (ie all stars are projected onto a sphere of infinite radius centered on the Earth). Two angles characterize the positions of objects on the celestial sphere: right ascension and declination. Right ascension, $\alpha$, is the angle measured eastwards along the celestial equator between the vernal equinox and the hour circle passing through an object.  Declination, $\delta$, is the angular distance of an object above or below the celestial equator.  
\par The heliocentric galactic coordinate system (often ambiguously referred to as galactic coordinates), is a different spherical coordinate system centered at the Sun.  The heliocentric fundamental plane is aligned with the galactic plane, and the primary direction is along the line connecting the sun and the galactic center.  Positions are measured in galactic latitude and longitude. Latitude, $b$, measures angular distance of an object perpendicular to the galactic plane, with positive angles to the north galactic pole. Longitude, $l$, is the angular distance measured eastward along the galactic equator from the galactic center.
\par The galactocentric coordinate system used in this paper is the same as the one used by van der Marel \citep{VdM08}.  The $Z$ axis points towards the galactic north pole, the $Y$ axis is oriented in the direction of the sun's rotation about the galactic center, and the $X$ axis points from the sun towards the galactic center. This is a standard, right-handed coordinate system.
\subsection{Equatorial to Galactic Transformation}
\label{sec:eq_to_gal}
\par This transformation is a rotation from equatorial coordinates to galactic coordinates. Given the right ascension and declination of a galaxy, $(\alpha,\delta)$, the transformation returns the galactic latitude and longitude, $(l,b)$. We use the same method described by \cite{poleski2013}, solving the following system of equations to find the galactic coordinates:
\begin{equation}
\begin{aligned}
\sin b &= \cos\delta\cos\delta_{G}\cos(\alpha - \alpha_{G}) + \sin\delta\sin\delta_{G} \\
\sin(l_{\Omega} - l)\cos b &= \cos\delta\sin(\alpha - \alpha_{G}) \\
\cos(l_{\Omega} - l)\cos b &= \sin\delta\cos\delta_{G} - \cos\delta\sin\delta_{G}\cos(\alpha - \alpha_{G})
\end{aligned}
\end{equation}

where $\alpha_{G} = 192.85948$ and $\delta_{G} = 27.12825$ are the equatorial coordinates of the galactic north pole, and $l_{\Omega}$ is the galactic longitude of the ascending node of the galactic plane.
\subsection{Galactic Spherical to Cartesian}
\label{sec:sphere_to_cart}
Before transforming to the galactocentric reference frame, we first transform the galactic latitude and longitude to cartesian coordinates. Expressing the coordinates in radians, the position in cartesians is simple:
\begin{equation}
\begin{aligned}
x &= D\cos b\sin l \\
y &= D\cos b\sin l \\
z &= D\sin b
\end{aligned}
\end{equation}
where $D$ is the distance to the observed galaxy (M31 or M33). 
\par The velocity transformation is more complicated.  They are given by:
\begin{equation}
\begin{aligned}
v_{x} &= v_{r}\cos(b)cos(l) - v_{W}\sin(l) - v_{N}\sin(b)\cos(l)\\
v_{y} &= v_{r}\cos(b)\sin(l) + v_{W}\cos(l) - v_{N}\sin(b)\sin(l) \\
v_{z} &= v_{r}\sin(b) + v_{N}\cos(b)
\end{aligned}
\end{equation}
where $v_{r}$ is the line of sight velocity of the observed galaxy, and $(v_{W},v_{N})$ are the linear velocities in the west and north directions as measured in the equatorial frame. In our project, we treat the velocity transformations of M31 and M33 differently due to better constraints on $(v_{W},v_{N})$ for M31 existing in the literature. For M31, we use van der Marel's constraints as the "observed" values: $(v_{W},v_{N}) = (-125.2 \pm 30.8, -73.8 \pm 28.4)$ \citep{vdm12m31}. There are only observations of the proper motion of M33 in spherical equatorial coordinates, so an extra transformation must be made to find the $(v_{W},v_{N})$ for M33. We follow van der Marel's method exactly from \cite{VdM08}.
\subsection{Heliocentric to Galactocentric}
\label{sec:hel_to_gal}
Transforming to the Galactocentric cartesian frame from the Heliocentric cartesian frame is a simple translation in the $X$ direction. The galactocentric positions are thus:
\begin{equation}
\begin{aligned}
x_{g} &= x_{h} - R_{0}
y_{g} &= y_{h} \\
z_{g} &= z_{h}
\end{aligned}
\end{equation}
where $R_{0} = 8.29 \pm 0.16$ kpc as used in \cite{vdm12m31}. The velocity transformations are slightly more complicated than a normal translation because the effects of the solar velocity in the central galaxy must be removed. We use the values derived by Schönrich et al: $v_{\odot} = (11.1, 12.24, 7.25)$ with the listed uncertainties \citep{sunlsr}. Thus the velocities in the galactocentric frame are:
\begin{equation}
\begin{aligned}
v_{x,g} &= v_{x,h} - v_{\odot,x} \\
v_{y,g} &= v_{y,h} - (V_{0}+v_{\odot,y})\\
v_{z,g} &= v_{z,h} - v_{\odot,z}
\end{aligned}
\end{equation}
where $V_{0} = 239.3 \pm 10.3$ is the circular velocity of the local standard of rest about the galactic center \citep{vdm12m31}.
% ----------------------------------------------------------------------------

@@

@@

\subsection{The \consuelo Simulation}
\label{sec:sampling_b}

@@

\subsection{The Local Group prior}
\label{sec:sampling_c}

@@

\begin{figure*}
  \centering \includegraphics[width=\textwidth]{plots/model_gof.png}
  \caption{The 1-D and 2-D marginalizations of the parameter space used to create the GMM model. The sample generated from the observed values of the 6 parameters (blue) and the sample drawn from the resulting GMM (red) overlap almost perfectly.}
  \label{fig:gof}
\end{figure*}

\begin{figure*}
  \centering \includegraphics[width=\textwidth]{plots/sim_plot.png}
  \caption{The 1-D and 2-D marginalizations of the CONSUELO triplet sample (green) overlayed with the corresponding marginalizations from the sample drawn from the GMM model (red). The intersection shows which of the CONSUELO triplets are given the most weight by the model.}
  \label{fig:sim_plot}
\end{figure*}

@@

\begin{figure}
\includegraphics[width=\linewidth]{plots/res_plot.png}
  \caption{Log masses (units of $M_{\odot}$) of the MW, M31, and the LG weighted by likelihood. The displayed values above the histograms are the triplet numbers.}
  \label{fig:res_plot}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{plots/pdf_lg_plot.png}
  \caption{The likelihood distribution of the mass of the Local Group. The Triplet distribution suffers from small statistics.}
  \label{fig:lg_pdf}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{plots/pdf_mw_plot.png}
  \caption{The likelihood distribution of the mass of the MW. The Triplet distribution suffers from small statistics.}
  \label{fig:mw_pdf}
\end{figure}
